---
title: "Ec 980FF Final Paper Summary"
author: "Buddy Scott"
date: "10/15/2020"
output: pdf_document
---

This paper will analyze the potential ramifications of the ongoing COVID-19
pandemic on the finances of the NBA and the market for its players. The NBA 
salary cap is the mechanism that effectively equilibrates supply and demand 
for players. The salary level is set based on Basketball Related Income (BRI).
NBA Commissioner Adam Silver revealed in May 2020 that ticket sales for
in-person fan attendances comprises an estimated 40% of the NBA's revenue. The 
NBA's revenue is roughly split 50-50 between players and the owners of these 
teams. Figure 1 shows how the players' and owners' split of BRI would decrease
according to this projection by Silver from $4 billion to $2.4 billion. If this
is the case, the league might have to withhold an increased percentage of 
players' salaries in escrow for the upcoming season to finance the league's
operations, potentially leaded to heavily contested debates at the bargaining
table of an updated Collective Bargaining Agreement. 

The two pressing economic problems this paper will address are the league's 
potentially inaccurate forward guidance provided to teams about salary cap 
levels for the upcoming season, as well as potential inequalities emerging
from teams' varying levels of cash flows and spending power that could 
potentially harm competitive balance in a zero-sum league.




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(gt)
```

```{r}

# Reading in the data
nbainfo <- read_csv("nbainfo.csv", 
                    col_type = cols(team = col_character(), 
                                    "1920winpct" = col_number(), 
                                    winpctrank = col_number(), 
                                    "1920ortg" = col_number(),
                                    ortgrank = col_number(), 
                                    "1920drtg" = col_number(), 
                                     drtgrank = col_number(), 
                                    "1920nrtg" = col_number(), 
                                     nrtgrank = col_number(), 
                                    "1920pace" = col_number(), 
                                     pacerank = col_number(), 
                                    gtcontracts = col_number(), 
                                    po_ngs = col_number(), 
                                    avgage = col_number(), 
                                    medage = col_number(), 
                                    avgexp = col_number(), 
                                    medexp = col_number(), 
                                    "2021projsalary" = col_number(), 
                                    "2021projcapspace" = col_number(), 
                                    "2021projexceptions" = col_character(), 
                                    tenyrwin = col_number(), 
                                    tenyrloss = col_number(), 
                                    tenyrwinpct = col_number(), 
                                    winpctrank_10 = col_number(), 
                                    playoffpct = col_number(), 
                                    principal_owner = col_character(), 
                                    nw = col_number(), 
                                    owned_since = col_number(), 
                                    business = col_character(), 
                                    other_owners = col_character(), 
                                    valuation = col_number(), 
                                    percent_change = col_number(), 
                                    debt_to_value = col_number(), 
                                    revenue = col_number(), 
                                    operating_income = col_number())) %>%
  rename_with(~ str_replace(.x, "1920", "current")) %>% 
  rename_with(~ str_replace(.x, "2021", "future")) %>% 
  slice(1:30) %>%
  subset(select = -futureprojcapspace)

playercontracts <- read_csv("bbrefcontractdata.csv", col_type = cols(
  playername = col_character(),
  playerid = col_character(),
  team = col_character(),
  salary1920 = col_double(),
  salary2021 = col_double(),
  salary2122 = col_double(),
  salary2223 = col_double(),
  salary2324 = col_double(),
  salary2425 = col_double(),
  signedusing = col_character(),
  guaranteed = col_double()
))

playercontracts_modified <- playercontracts %>%
  filter(salary2021 >= 20000000) %>%
  arrange(desc(salary2021))
```

```{r}

tibble(subject = c("2019-2020 Season", 
                   "Projection for Next Season With No Fans"),
       ynear = c("$8B", "$4.8B"),
       yfar = c("$4B", "$2.4B"), 
       ydiff = c("$4B", "$2.4B")) %>%
    
    # table setup
    gt() %>%
    cols_label(subject = "BRI",
               ynear = "Total",
               yfar = "Players",
               ydiff = "Owners") %>%
    tab_style(cell_borders(sides = "right"),
              location = cells_body(columns = vars(subject))) %>%
    tab_style(cell_text(weight = "bold"),
              location = cells_body(columns = vars(subject))) %>%
    cols_align(align = "center", columns = TRUE) %>%
    fmt_markdown(columns = TRUE) %>%
    tab_header(title = "Figure 1: Potential BRI Drop in 2020-2021 Season", 
               subtitle = "Commisssioner Adam Silver projects ticket sales
               comprise 40% of BRI")
```

```{r}
# Some cool graphs

nbainfo %>%
ggplot(aes(x = currentwinpct, y = valuation, label = team)) + 
  geom_point() + 
  stat_smooth(formula = y ~ x, method = "lm", se = TRUE) + 
  geom_text(nudge_x = 0.02) +
  annotate(x=0.5, y=3.5, label=paste("Correlation = ", 
                                     round(cor(nbainfo$currentwinpct, 
                                               nbainfo$valuation), 3)), 
           geom="text", size=5) + 
  labs(title = "Figure 2: Team Valuation vs. 2019-2020 Winning Percentage",
       x = "2019-2020 Winning Percentage", y = "Team Valuation") + 
  scale_y_continuous(breaks = c(0, 1, 2, 3, 4),
                     labels = c("$0", "$1B", "$2B", "$3B", "$4B"))

nbainfo %>%
ggplot(aes(x = tenyrwinpct, y = valuation, label = team)) + 
  geom_point() + 
  geom_text(nudge_x = 0.01) + 
  stat_smooth(formula = y ~ x, method = "lm", se = TRUE) + 
  annotate(x=0.5, y=3.5, label=paste("Correlation = ", 
                                     round(cor(nbainfo$tenyrwinpct, 
                                               nbainfo$valuation), 3)), 
         geom="text", size=5) + 
  labs(title = "Figure 3: Team Valuation vs. Ten Year Winning Percentage",
       x = "Ten Year Winning Percentage", y = "Team Valuation") + 
  scale_y_continuous(breaks = c(0, 1, 2, 3, 4),
                     labels = c("$0", "$1B", "$2B", "$3B", "$4B"))

nbainfo %>%
  ggplot(aes(x = fct_reorder(team, valuation), y = valuation)) + 
    geom_col() + 
    coord_flip() + 
    labs(title = "Figure 4: Team Valuations", x = "February 2020 Valuation", 
         y = "Team") + 
    scale_y_continuous(breaks = c(0, 1, 2, 3, 4),
                     labels = c("$0", "$1B", "$2B", "$3B", "$4B"))

nbainfo %>%
  ggplot(aes(x = fct_reorder(team, nw), y = nw)) + 
    geom_col() + 
    coord_flip() + 
    labs(title = "Figure 5: Team Majority Owner's Net Worth", x = "Net Worth", 
         y = "Team") + 
    scale_y_continuous(breaks = c(0, 20, 40, 60, 80),
                     labels = c("$0", "$20B", "$40B", "$60B", "$80B"))
  
  


```

\newpage
## Code

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```